---
output:
  md_document:
    variant: gfm
always_allow_html: true
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

# NOAAsignifEarthQuakes

The goal of NOAAsignifEarthQuakes is to:

1. Read and clean data from the NOAA Significant Earthquake Database^[Source: National Geophysical Data Center / World Data Service (NGDC/WDS); Significant Earthquake Database; National Geophysical Data Center, NOAA. doi:10.7289/V5TD9V7K.]
1. Provide exploratory visualization tools:
    1. Timelines of earthquakes by countries
    1. Mapping of earthquakes in a specific region

## Example

### Loading the package

```{r dependencies}
library(NOAAsignifEarthQuakes)
library(magrittr)
```

### Reading and cleaning the data

```{r read_and_clean}
## set the file input
file_noaa <- system.file("extdata","signif.txt",
                         package="NOAAsignifEarthQuakes",
                         mustWork=TRUE)
## Read and clean data
noaa <- load_NOAA_db(file_noaa) %>% eq_clean_data()
knitr::kable(tail(noaa))
```

### Display a timeline

```{r timeline_usa}
geom_timeline(noaa,
              mapping=ggplot2::aes(size=MAG,fill=DEATHS),
              countries='USA',
              xmin='2000-01-01',
              xmax='2017-01-01')
```

### Display a timeline with labels

```{r timeline_usa_china_labels}
g <- geom_timeline(noaa,
                   mapping=ggplot2::aes(size=MAG,fill=DEATHS,y=COUNTRY),
                   countries=c('USA','China'),
                   xmin='2000-01-01',
                   xmax='2017-01-01') 
g + geom_timeline_label()
```


### Display map of Earthquakes locations

```{r map_mexico}
filt_noaa <- noaa %>% dplyr::filter(COUNTRY == "MEXICO" & lubridate::year(DATE) >= 2000)
m <- eq_map(filt_noaa,annot_col = "popup_text")
```

The map is fully interactive, for documentation purposes we show the following static version.

```{r static_display, echo=FALSE, fig.width=10, fig.height=8}
map_png <- "README-map_mexico-1.png"
mapview::mapshot(m, file = map_png)
knitr::include_graphics(map_png)
```
